---
title: "Introduction to R. iMOBAs 2022"
author: "BioinfMOBA"
format: revealjs
---

## Arithmetic Operations

There are several important arithmetic operations defined in R: 

```{r}
#| echo: true
#| eval: false

+ x
- x
x + y 
x - y 
x * y
x / y
x^y
x %% y
x %/% y
```

**Q**: In your *console* type this expression. What is the result?

```{r}
#| echo: true
#| eval: false
#| label: operators_usage

(- 1 + (29 - 58)) * (1 / 30) / -9999.921^0
```
**Q**: Arithmetic operations are *unary* and *binary*. What are those?

## Variables

Think about the variable as a cell, or basket *where you store the data*. Well, really it's a *name* you choose to call a *memory slot* where your data stored. *Pointer* if it makes a point?

We can perform all the mentioned arithmetic operations on variables, as we did with numbers above, and *store* the results to *the same* or *other* variables

```{r}
#| echo: true
#| eval: false

x <- 5  %% 3
y <- x
x <- (x + y)^3
y <- y - 1
z <- x * y
```

**Q**: What is the value of `z`?

## Vectors: Introduction

Think about the *vectors* as a contiguous cells for storing the data. 

Important note: all the elements of the vector must be of the same *class*.

```{r}
#| echo: true
#| eval: false
apples <- c('ðŸŽ', 'ðŸŽ', 'ðŸŽ')
oanges <- c('ðŸ‹', 'ðŸ‹', 'ðŸ‹')
```

You can't compare apples with oranges. This is why you can't store different types in the same vector!

**Q**: Using the syntax for fruits, make a vector of few `\U1F680` elements.

## Vectors: Indexing

Each element of the vector, can be accessed by the *index* pointing to the *location* of the element. Try it by your own:

```{r}
#| echo: true
#| eval: false
fbn <- c(1, 1, 2, 3, 5, 8, 13)

fbn[4]
fbn[2:5]
fbn[7:1]
fbn <- fbn[-2]
```

**Q**: The murderer terrorizes your neighborhood. Every night going out, you are checking the content of your handbag. There is one thing which definitely won't help you to defend yourself. Remove it from your handbag.

```{r}
#| echo: true
#| eval: false
handbag <- c('\U1F5E1', '\U1F528', '\U1F36D', '\U1F4A3')
```

::: footer
R uses *one-based* indexing, i.e. we start counting from `1`.
:::

## Vectors: Em ain't not for numbers only!

We can store other things in vectors: `character`s, `boolean`s, `complex` numbers, `function`s - you name it!

```{r}
#| echo: true
#| eval: false

nums  <- c(1, -1, 1, -1, 1, -1)
chars <- c('a', 'b', 'c', '\U1F353', 'I am a string')
bools <- c(TRUE, FALSE, TRUE, TRUE, TRUE == FALSE)
```

**Q**: What is the fourth entry of `chars`? 

## Vectors: length, c

Vector is an *object*, or simply a data structure with *methods* associated with it. The `length()` method returns the *number of elements* in the vector:

```{r}
#| echo: true
#| eval: false
set.seed(0)
x <- vector(length = sample(100000, 1))
```

**Q**: What is the length of this vector?

The name of `c()` method is a *shorthand* for *concatenate* - it tries to *combine* all it's arguments into one vector:

```{r}
#| echo: true
#| eval: false
bools  <- c(TRUE, FALSE)
nums   <- c(0, 1)
whoami <- c(nums, bools)
```

**Q**: What is the result of the last line? Why we got these results? 

## Vectors: Vectorized operations

*Most* of the operations and functions in R are *vectorized*. It simply means that unlike other languages, you don't need to iterate over each element of the vector to apply the same operation over it. You can do it at once, on the entire vector:

```{r}
#| echo: true
#| eval: false
x <- 1:100
x - 1
x * 2
x^2
```

**Q**: Execute expression `x - x`, explain the result. Execute the expression `x - x[1:50]` can you explain the result? Sum two subsets of `x`, i.e. execute `x[1:49] + x[99:51]`. Examining the result can you find the sum of all elements of `x`?

## Vectors: Useful functions on vectors

There are many useful *functions* shipped with Base R, which might be applied to the vectors. Below is only the small subset of them:

```{r}
#| echo: true
#| eval: false
set.seed(0)
x <- sample(100, 10, TRUE)
sum(x)    # sum of all elements
mean(x)   # expected value of x
median(x) # median of x
```
You can find detailed description of the functions listed above (and any other function!) with `?<function name>` syntax.

## Matrix

- *Matrix* is just an *array* of vectors! 
- A `matrix` contains vectors of the same length;
- A `matrix` is *homogeneous* - it contains data of the *same* type;
- A `matrix` has *rows* and *columns*, with rows and columns *names* and *indices* associated with them;
- All the operations / functions defined on vectors are defined on matrices and more...;

And remember nearly *everything* in R is *vectorized* ðŸš€.

## Matrix 

Run an example

```{r}
#| echo: true
#| eval: false
set.seed(0)
m <- matrix(
    c(2, 7, 6, 9, 5, 1, 4, 3, 8),
    ncol = 3,
    byrow = TRUE
)

nrow(m)
ncol(m)
```

**Q**: What is the `sum()` of all the matrix elements? 

**Q**: Run the code below in your console. Explain the result.

```{r}
#| echo: true
#| eval: false
m[1, 1] <- '\U1F575'
rowSums(m)
```


## List

- `list` can contain *different* data types of *different length*;
- `list` can be *empty*, *flat* or *nested* and the values can be accessed by *indices / names*;
- Functions and operations in R *generally* are not vectorized over the lists$^1$! 

```{r}
#| echo: false
#| eval: true
lst <- list(
    fruits  = c(orange = '\U1F34A', lemon = '\U1F34B', banana = '\U1F34C'),
    drinks  = c(wine = '\U1F377', milk = '\U1F95B'),
    wallet  = list(
        '\U20AC' = 24,
        '\U0024' = 8,
        '\U20B4' = 1991
    )
)

dplyr::glimpse(lst)
```

:::: footer
$^1$You have to *iterate* over them - this is where functions from `-apply` family shine ðŸš€!
::::

## `data.frame`

- A `data.frame` is a `list` of vectors of *equal length*;
- Think about it as a *table* with rows (*samples / trials*) and columns (*variables*);
- Anything applicable to the *vectors* and *lists* is applicable to the `data.frame`;

```{r}
#| echo: true
#| eval: true

data.frame(
    fruits = lst$fruits,
    money  = sapply(lst$wallet, I),
    unit   = names(lst$wallet)
)
```

## Iris data

1. Load the `iris` data into your environment;
2. Use one of the functions explained above, to find:
  1.1 Number of `sentosa` species in the data set;
  1.2 Max `Petal.Length` in the data set;
  1.3 Average `Sepal.Width` in the data set;
3. Call `boxplot(iris)` and find what is wrong with this representation;

## Iris data

Load the data:
```{r}
#| echo: true
#| eval: true
data(iris)
```

Use the `summary` built-in function, to get the summary statistics:
```{r}
#| echo: true
#| eval: true
summary(iris)
```

## Iris data

```{r}
par(
    lwd = .5,
    col.axis = 'gray25',
    col.lab  = 'gray25',
    col.main = 'gray25',
    col.sub = 'gray50',
    family  = 'mono',
    pty = 's',
    tcl = -.1
)
```

Plot the data:
```{r}
#| echo: true
#| eval: true
boxplot(
    iris,
    col  = 'white',
    notch = TRUE
)
```

## Part II task 

1. Attach a `DNAse` data;
2. Convert `DNAse` data set from wide to long format. Use `conc` and `density` as a `measure.vars`;
3. Add new columns calculating the `median()` and `log2()` for `Run` and `variable`;
4. All the new object `DNAse_new`;

## Part III theme() trick

You can set your theme at the begining of the script, so you don't have to call everytime you make a plot:

```{r}
#| echo: true
#| eval: false
theme_set(
  ggthemes::thme_fiew()
)
```

## Part III Tasks

Make plots using: [https://r-graph-gallery.com/](https://r-graph-gallery.com/), [https://r-charts.com/](https://r-charts.com/).

1. Plot 1: `iris` using exaple from `r-charts.com`;
2. Plot 2: `iris` using example from `r-graph-gallery.com`;
3. Plot 3: `iris` data set, bake your own graph!
4. Plot 4: `french_fries` data set using example from [https://r-graph-gallery.com/](https://r-graph-gallery.com/);
5. Plot 5: `DNAse` data set using example from [https://r-graph-gallery.com/](https://r-graph-gallery.com/);

